---
title: "Exploratory conditional whole-MAG DNA abundance–phenotype associations"
format: html
editor: source
---

Author-approved exploratory modelling of digitised published means.14 MAGs x3
outcomes, one model per pair. This is conditional association, not phage efficacy.
Model: transformed outcome ~ phase + factor(cycle) + condition + log2_MAG_abundance.
Cells retain published log10 units; EPS endpoints are log10 transformed.
Expression selects candidates only. Predictor is log2 whole-MAG DNA counts divided
by frozen DESeq2 size factors; not absolute cell counts or gene expression.
MAGs were selected post hoc using upregulated genes, not independent validation.
Condition is confounded with membrane;12 observations and6 residual df per fit.
Nominal OLS and HC3 t tests with residual df, separate BH families of42 each.
Neither provides design-based repeated-membrane causal inference.
Leave-one-sample and leave-one-cycle refits are influence checks, not independent
cross-validation. Pixel envelopes are extraction bounds, not sampling intervals.
No TMP duplication across phases, no composite or technical pseudo-replicates.

```{r}
suppressPackageStartupMessages({library(data.table);library(ggplot2)})
out<-commandArgs(TRUE)[1];stopifnot(!dir.exists(out))
root<-"/Users/shaman.narayanasamy/Work/data/phage_uv_treatment/PRJEB79569/derived"
src<-file.path(root,"figure_derived_biofilm_phenotypes_2026-09-13-v1/tables/expression_phenotype_overlay.tsv")
sha<-function(p)digest::digest(file=p,algo="sha256")
before<-sha(src);z<-fread(src)
dna_dir<-file.path(root,"community_cycle_models_2026-09-13-v1/tables")
dna_path<-file.path(dna_dir,"mag_normalized_counts.tsv")
raw_path<-file.path(dna_dir,"mag_counts.tsv");sf_path<-file.path(dna_dir,"mag_size_factors.tsv")
dna<-fread(dna_path);raw<-fread(raw_path);sf<-fread(sf_path)
stopifnot(identical(dna$feature_id,raw$feature_id))
for(sn in sf$sample_title)stopifnot(max(abs(dna[[sn]]-raw[[sn]]/sf[sample_title==sn,size_factor]))<1e-6)
long<-melt(dna,id.vars="feature_id",variable.name="sample_title",value.name="normalized_DNA_count")
setnames(long,"feature_id","MAG_ID")
z[,expression_score:=NULL]
z<-merge(z,long,by=c("MAG_ID","sample_title"),all.x=TRUE)
stopifnot(!anyNA(z$normalized_DNA_count),all(z$normalized_DNA_count>0))
z[,log2_MAG_abundance:=log2(normalized_DNA_count)]
stopifnot(nrow(z)==504,!anyDuplicated(z[,.(MAG_ID,outcome,sample_title)]))
for(d in c("tables","figures","source"))dir.create(file.path(out,d),recursive=TRUE)
ans<-list();influence<-list();residuals<-list();fits<-list()
for(mag in sort(unique(z$MAG_ID)))for(o in sort(unique(z$outcome))){
 d<-copy(z[MAG_ID==mag & outcome==o]);setorder(d,sample_title)
 d[,`:=`(phase=factor(phase),cycle=factor(cycle),condition=factor(condition))]
 stopifnot(nrow(d)==12,all(d$pixel_envelope_low>0))
 transform_y<-function(y)if(o=="log10_cells")y else log10(y)
 d[,`:=`(y=transform_y(estimate),ylow=transform_y(pixel_envelope_low),yhigh=transform_y(pixel_envelope_high))]
 fit<-lm(y~phase+cycle+condition+log2_MAG_abundance,data=d)
 X<-model.matrix(fit);stopifnot(qr(X)$rank==ncol(X),df.residual(fit)==6)
 j<-match("log2_MAG_abundance",colnames(X));V<-solve(crossprod(X));weights<-V%*%t(X)
 beta<-coef(fit)[j];se<-sqrt(diag(vcov(fit)))[j];df<-df.residual(fit)
 hc3<-V%*%crossprod(X,X*as.numeric((resid(fit)/(1-hatvalues(fit)))^2))%*%V
 se3<-sqrt(diag(hc3))[j]
 # Exact coefficient range over independent transformed-y extraction envelopes.
 w<-weights[j,];lo<-sum(ifelse(w>=0,w*d$ylow,w*d$yhigh));hi<-sum(ifelse(w>=0,w*d$yhigh,w*d$ylow))
 nuisance<-lm(log2_MAG_abundance~phase+cycle+condition,data=d)
 inf<-rbindlist(lapply(seq_len(nrow(d)),function(i){
  dd<-droplevels(d[-i]);ff<-lm(y~phase+cycle+condition+log2_MAG_abundance,data=dd)
  data.table(drop_type="sample",dropped=as.character(d$sample_title[i]),beta=coef(ff)['log2_MAG_abundance'],rank=ff$rank)
 }))
 inf<-rbind(inf,rbindlist(lapply(levels(d$cycle),function(cy){
  dd<-droplevels(d[cycle!=cy]);ff<-lm(y~phase+cycle+condition+log2_MAG_abundance,data=dd)
  data.table(drop_type="cycle",dropped=cy,beta=coef(ff)['log2_MAG_abundance'],rank=ff$rank)
 })))
 inf[,`:=`(MAG_ID=mag,outcome=o,same_sign=sign(beta)==sign(as.numeric(coef(fit)[j])))]
 influence[[paste(mag,o)]]<-inf
 ans[[paste(mag,o)]]<-data.table(MAG_ID=mag,genus=d$genus[1],outcome=o,n=12,df=df,
  beta=beta,se=se,CI_low=beta-qt(.975,df)*se,CI_high=beta+qt(.975,df)*se,
  p=2*pt(-abs(beta/se),df),HC3_se=se3,HC3_p=2*pt(-abs(beta/se3),df),
  partial_R2=(beta/se)^2/((beta/se)^2+df),predictor_VIF=1/(1-summary(nuisance)$r.squared),
  design_kappa=kappa(X),max_Cook=max(cooks.distance(fit)),max_leverage=max(hatvalues(fit)),
  pixel_beta_low=lo,pixel_beta_high=hi,pixel_direction_stable=lo*hi>0,
  leave_sample_same_sign=sum(inf[drop_type=="sample",same_sign],na.rm=TRUE),
  leave_cycle_same_sign=sum(inf[drop_type=="cycle",same_sign],na.rm=TRUE))
 d[,`:=`(fitted=fitted(fit),residual=resid(fit),leverage=hatvalues(fit),Cook=cooks.distance(fit),
  partial_abundance=resid(nuisance),partial_y=resid(lm(y~phase+cycle+condition,data=d)))]
 residuals[[paste(mag,o)]]<-d
 fits[[paste(mag,o)]]<-fit
}
a<-rbindlist(ans);stopifnot(nrow(a)==42)
a[,`:=`(FDR=p.adjust(p,"BH",n=42),HC3_FDR=p.adjust(HC3_p,"BH",n=42))]
fwrite(a,file.path(out,"tables/model_results.tsv"),sep="\t")
fwrite(rbindlist(influence),file.path(out,"tables/deletion_sensitivity.tsv"),sep="\t")
fwrite(rbindlist(residuals),file.path(out,"tables/model_observations.tsv"),sep="\t")
saveRDS(fits,file.path(out,"fitted_models.rds"))
```

```{r}
#| label: verify-and-preview
# Frisch-Waugh-Lovell independent slope check for all42fits.
rr<-rbindlist(residuals)
check<-rr[,.(beta_FWL=sum(partial_abundance*partial_y)/sum(partial_abundance^2)),by=.(MAG_ID,outcome)]
check<-merge(check,a[,.(MAG_ID,outcome,beta)],by=c("MAG_ID","outcome"))
stopifnot(max(abs(check$beta-check$beta_FWL))<1e-9,all(a$pixel_beta_low<=a$beta),all(a$pixel_beta_high>=a$beta))
# End-point construction verifies the analytic pixel bounds directly.
for(key in names(fits)){
 ff<-fits[[key]];dd<-residuals[[key]];xx<-model.matrix(ff)
 w<-solve(crossprod(xx),t(xx))["log2_MAG_abundance",]
 low_y<-ifelse(w>=0,dd$ylow,dd$yhigh)
 high_y<-ifelse(w>=0,dd$yhigh,dd$ylow)
 stopifnot(abs(coef(lm.fit(xx,low_y))["log2_MAG_abundance"]-sum(w*low_y))<1e-9,
  abs(coef(lm.fit(xx,high_y))["log2_MAG_abundance"]-sum(w*high_y))<1e-9)
}
a[,label:=paste(ifelse(is.na(genus)|genus=="","Unclassified",genus),MAG_ID,sep=" / ")]
a[,nominal_support:=ifelse(FDR<.05,"OLS FDR < 0.05","OLS FDR >= 0.05")]
g<-ggplot(a,aes(beta,label,colour=nominal_support))+geom_vline(xintercept=0,colour="grey65")+
 geom_segment(aes(x=CI_low,xend=CI_high,yend=label),linewidth=.5)+geom_point(size=2)+
 facet_wrap(~outcome,nrow=1,scales="free_x")+theme_bw(base_size=9)+
 scale_colour_manual(values=c("OLS FDR < 0.05"="#175D83","OLS FDR >= 0.05"="#777777"))+
 labs(x="Conditional slope (log10 outcome per doubling of whole-MAG DNA abundance)",y=NULL,
  subtitle="Figure-derived outcomes; adjusted for phase, cycle and condition; nominal OLS 95% intervals",
  caption="Exploratory, post hoc candidates. Intervals assume independent model errors; not phage-efficacy predictions.")+theme(legend.position="bottom")
ggsave(file.path(out,"figures/conditional_associations.png"),g,width=16,height=8,dpi=160,device=ragg::agg_png)

# Fixed-normalization suppression scenarios: same fitted model, lower whole-MAG DNA.
sc<-rbindlist(lapply(seq_len(nrow(a)),function(i){
 aa<-a[i];dd<-rr[MAG_ID==aa$MAG_ID & outcome==aa$outcome]
 ss<-data.table(suppression=c(0,seq(.1,.9,.1),.95))
 ss[,delta_log2:=log2(1-suppression)]
 ss[,`:=`(MAG_ID=aa$MAG_ID,genus=aa$genus,outcome=aa$outcome,
  change_pct=100*(10^(aa$beta*delta_log2)-1),
  OLS_low_pct=100*(10^(aa$CI_high*delta_log2)-1),
  OLS_high_pct=100*(10^(aa$CI_low*delta_log2)-1),
  HC3_low_pct=100*(10^((aa$beta+qt(.975,aa$df)*aa$HC3_se)*delta_log2)-1),
  HC3_high_pct=100*(10^((aa$beta-qt(.975,aa$df)*aa$HC3_se)*delta_log2)-1),
  FDR=aa$FDR,HC3_FDR=aa$HC3_FDR)]
 ss[,treatment_samples_below_observed_range:=vapply(delta_log2,function(dx)
  sum(dd[condition=="treatment",log2_MAG_abundance]+dx<min(dd$log2_MAG_abundance)),integer(1))]
 ss
}))
stopifnot(nrow(sc)==462,all(sc[suppression==0,change_pct]==0))
# Independently check coefficient contrasts with predict() at 50% suppression.
for(key in names(fits)){
 dd<-copy(residuals[[key]]);base<-predict(fits[[key]],newdata=dd)
 dd[,log2_MAG_abundance:=log2_MAG_abundance-1]
 stopifnot(max(abs(predict(fits[[key]],newdata=dd)-base+coef(fits[[key]])["log2_MAG_abundance"]))<1e-9)
}
fwrite(sc,file.path(out,"tables/abundance_suppression_scenarios.tsv"),sep="\t")
proxy<-rbindlist(lapply(c(.5,.75,1,1.25,1.5),function(rho){
 dd<-copy(sc);dd[,efficiency_ratio:=rho]
 dd[,burden_proxy_change_pct:=100*((1+change_pct/100)/rho-1)]
 dd
}))
fwrite(proxy,file.path(out,"tables/detachment_efficiency_sensitivity.tsv"),sep="\t")
g2<-ggplot(sc[suppression==.5],aes(change_pct,paste(genus,MAG_ID,sep=" / ")))+
 geom_vline(xintercept=0,colour="grey65")+geom_segment(aes(x=OLS_low_pct,xend=OLS_high_pct,yend=paste(genus,MAG_ID,sep=" / ")),linewidth=.4)+
 geom_point(colour="#175D83",size=2)+facet_wrap(~outcome,nrow=1,scales="free_x")+
 theme_bw(base_size=9)+labs(x="Predicted proxy change at 50% lower whole-MAG DNA abundance (%)",y=NULL,
 subtitle="Fixed-normalization scenario; nominal OLS 95% intervals",
 caption="Expression-selected MAGs; separate conditional models. Not measured phage efficacy.")
ggsave(file.path(out,"figures/abundance_suppression50.png"),g2,width=16,height=8,dpi=180,device=ragg::agg_png)

stopifnot(identical(before,sha(src)))
notebook<-Sys.getenv("PHAGE_UV_NOTEBOOK_PATH")
file.copy(notebook,file.path(out,"source",basename(notebook)))
fwrite(data.table(path=c(src,dna_path,raw_path,sf_path,notebook),sha256=vapply(c(src,dna_path,raw_path,sf_path,notebook),sha,character(1))),file.path(out,"input_sha256.tsv"),sep="\t")
capture.output(sessionInfo(),file=file.path(out,"sessionInfo.txt"))
writeLines(c("PASS: DNA raw/size-factor reconciliation; positive candidate counts; 462 suppression scenarios; predict contrast checks.",
 "PASS:42full-rank models;6 residual df;independent FWL slopes;analytic pixel-bound verification.",
 "42OLS and42HC3 p-values adjusted separately; no sample independence guaranteed by HC3.",
 "Provisional published-cycle alignment and Backflush fraction ambiguity retained.",
 "Whole-MAG DNA normalized counts are the predictor; expression only selects candidates.",
 "Delete-one-cycle is sensitivity, not independent predictive validation."),file.path(out,"CHECKS.txt"))
files<-list.files(out,recursive=TRUE,full.names=TRUE)
fwrite(data.table(path=substring(files,nchar(out)+2),sha256=vapply(files,sha,character(1))),file.path(out,"output_sha256.tsv"),sep="\t")
```

