---
title: "MAG expression groups and hypothetical abundance reduction"
format: html
execute:
  enabled: false
---

## Specification fixed before fitting

The primary grouping compares the original 14 up-hit candidates
versus the remaining 334 MAGs. Secondary view distinguishes up-hit, down-only,
no significant main response, and no tested active-list genes. A joint cycle
interaction does not alter the main-effect label; retain it as a separate flag.

Display all individual-MAG 50% reduction estimates across the three existing
adjustments and both phases. Distribution summaries describe this finite
catalogue, not independent replicates. Do not compare 14 and 334 fitted slopes
with a naive t-test, Wilcoxon test, or independent bootstrap.

For a direct model contrast, define each group's abundance as the geometric
mean of member MAG normalized DNA counts (mean log2 counts). This gives each
MAG equal weight. A coordinated fractional reduction f in every member shifts
the group's predictor by log2(1-f), while other groups stay fixed. Fit group
predictors jointly to the six outcomes, separately for each phase. No summation
of coefficients from separate single-MAG models.

Binary joint model: score ~ condition + x_up + x_rest (df2); sensitivity adds
linear cycle (df1). Adding categorical cycle makes this saturated (df0), so
report that design audit without p-values or fitted intervention claims.
Directional joint model: score ~ condition + x_up + x_down + x_ns (df1), using
14/22/248 members, respectively. The 64 MAGs without tested genes are excluded
from these three predictors and remain a displayed fourth evidence category.
Adding cycle is saturated and unsupported for inferential fitting.

Primary comparison statistic: at 50% coordinated reduction, score change for
rest minus score change for up = beta_up - beta_rest. Two-sided t contrast from
the joint model covariance. Secondary directional comparisons: down minus up,
ns minus up, ns minus down. Ten planned contrasts total: binary 2 phases x 2
adjustments plus directional 2 phases x 3 comparisons. BH over all ten. Also
report HC3 sensitivity with separate BH10. All tests assume independent Gaussian
errors conditional on the fixed predictors. This is an explicitly exploratory
extension after the single-MAG results, not a preregistered analysis.

Export pointwise scenario intervals over 0%,10%,25%,50%,75%,90%,95%, leave-cycle-
out contrast estimates where full rank, joint predictor VIFs, leverage, condition
number and extrapolation flags. The 50% comparison is the same linear contrast
at other nonzero fractions, not seven independent tests. Equal percentages do
not mean equal numbers of taxa or equal biomass removed.

## Limitations

Only two membranes and three cycles inform these models. Group aggregation
changes the estimand from reducing one MAG to coordinated suppression of a group.
The geometric mean is an equal-MAG abundance index, not total community biomass.
Normalisation stays fixed. There is no calibrated repeated-membrane treatment
inference; one or two residual degrees of freedom yield fragile intervals.
Group membership was selected from expression in the same experiment. Counts,
score-source alignment and uncertainty limitations from the original models apply.

```{r}
suppressPackageStartupMessages({library(data.table);library(ggplot2)})
root <- "/Users/shaman.narayanasamy/Work/data/phage_uv_treatment/PRJEB79569/derived"
out <- commandArgs(TRUE)[1];stopifnot(length(out)==1,!is.na(out),!dir.exists(out))
paths <- c(evidence=file.path(root,"cleaning_score_all_MAG_models_2026-09-21-v1/tables/MAG_expression_evidence.tsv"),
 single=file.path(root,"cleaning_score_all_MAG_models_2026-09-21-v1/tables/50_percent_reduction_annotated.tsv"),
 dna=file.path(root,"community_cycle_models_2026-09-13-v1/tables/mag_normalized_counts.tsv"),
 meta=file.path(root,"community_cycle_models_2026-09-13-v1/tables/sample_metadata.tsv"),
 score=file.path(root,"cleaning_performance_score_2026-09-14-v1/scores.tsv"))
sha <- function(p)digest::digest(file=p,algo="sha256")
before <- vapply(paths,sha,character(1))
e <- fread(paths['evidence'])
e[,binary:=ifelse(existing_14_candidate,"up14","rest334")]
e[,direction:=fcase(upregulated_genes>0,"up",downregulated_genes>0,"down",tested_biofilm_genes>0,"ns",default="not_tested")]
stopifnot(nrow(e)==348,sum(e$direction=="up")==14,sum(e$direction=="down")==22,
 sum(e$direction=="ns")==248,sum(e$direction=="not_tested")==64,
 !any(e$upregulated_genes>0 & e$downregulated_genes>0))
for(dr in c("tables","figures","source"))dir.create(file.path(out,dr),recursive=TRUE)
fwrite(e,file.path(out,"tables/MAG_group_membership.tsv"),sep="\t")
s <- merge(fread(paths['single']),e[,.(MAG_ID,binary,direction)],by="MAG_ID")
summary <- rbindlist(lapply(c("binary","direction"),function(g){
 ans <- s[,.(MAGs=.N,median_gain=median(delta_score),mean_gain=mean(delta_score),
 q25=quantile(delta_score,.25),q75=quantile(delta_score,.75),minimum=min(delta_score),maximum=max(delta_score),
 positive_estimates=sum(delta_score>0),nominal_positive_interval=sum(low>0),BH_positive=sum(low>0 & FDR_696<.05),
 below_range_any=sum(treated_below_observed_range>0)),by=c("specification","phase",g)]
 setnames(ans,g,"group");ans[,grouping:=g];ans
}))
fwrite(summary,file.path(out,"tables/individual_MAG_group_summary.tsv"),sep="\t")
fwrite(s,file.path(out,"tables/individual_MAG_50_percent_by_group.tsv"),sep="\t")
dna <- fread(paths['dna']); long <- melt(dna,id.vars="feature_id",variable.name="sample_title",value.name="count")
setnames(long,"feature_id","MAG_ID");long <- merge(long,e[,.(MAG_ID,binary,direction)],by="MAG_ID")
stopifnot(all(long$count>0))
long[,log_count:=log2(count)]
aggregates <- rbindlist(lapply(c("binary","direction"),function(g){
 d <- long[,.(x=mean(log_count),members=.N),by=c("sample_title",g)];setnames(d,g,"group");d[,grouping:=g];d
}))
meta <- fread(paths['meta']);score <- fread(paths['score'])
aggregates <- merge(aggregates,meta[,.(sample_title,condition,phase,cycle)],by="sample_title")
aggregates <- merge(aggregates,score[,.(condition,cycle,score)],by=c("condition","cycle"))
fwrite(aggregates,file.path(out,"tables/group_abundance_observations.tsv"),sep="\t")
# Per-group contrast from a single joint covariance matrix.
contrast <- function(fit,L,V=vcov(fit)){
 b <- sum(L*coef(fit)); se <- sqrt(drop(t(L)%*%V%*%L)); df <- df.residual(fit)
 data.table(estimate=b,se=se,t=b/se,df=df,p=2*pt(-abs(b/se),df),low=b-qt(.975,df)*se,high=b+qt(.975,df)*se)
}
audit <- list();fits <- list();contrasts <- list();scenarios <- list();deletions <- list();diagnostics <- list();observations <- list()
for(g in c("binary","direction"))for(ph in c("initial","backflush")){
 d <- dcast(aggregates[grouping==g & phase==ph],sample_title+condition+cycle+score~group,value.var="x")
 d[,condition:=factor(condition,levels=c("control","treatment"))];setorder(d,condition,cycle)
 predictors <- if(g=="binary")c("up14","rest334") else c("up","down","ns")
 comparisons <- if(g=="binary")list(c("rest334","up14")) else list(c("down","up"),c("ns","up"),c("ns","down"))
 specs <- if(g=="binary")c("condition_only","linear_cycle","categorical_cycle") else c("condition_only","linear_cycle","categorical_cycle")
 for(spec in specs){
  nuisance <- switch(spec,condition_only="condition",linear_cycle="condition + cycle",categorical_cycle="condition + factor(cycle)")
  form <- as.formula(paste("score ~",nuisance,"+",paste(predictors,collapse=" + ")))
  fit <- lm(form,d);X <- model.matrix(fit);df <- df.residual(fit);key <- paste(g,ph,spec,sep="/")
  valid <- fit$rank==ncol(X) && df>0
  audit[[key]] <- data.table(grouping=g,phase=ph,specification=spec,n=6,parameters=ncol(X),rank=fit$rank,residual_df=df,inference_available=valid)
  if(!valid)next
  fits[[key]] <- fit; V <- vcov(fit);h <- hatvalues(fit);B <- solve(crossprod(X))
  V3 <- B%*%crossprod(X,X*as.numeric((resid(fit)/(1-h))^2))%*%B
  observations[[key]] <- cbind(data.table(grouping=g,phase=ph,specification=spec),d,fitted=fitted(fit),residual=resid(fit))
  for(pr in predictors){
   xf <- as.formula(paste(pr,"~",nuisance,"+",paste(setdiff(predictors,pr),collapse=" + ")))
   diagnostics[[paste(key,pr)]] <- data.table(grouping=g,phase=ph,specification=spec,group=pr,
    beta=coef(fit)[pr],VIF=1/(1-summary(lm(xf,d))$r.squared),max_leverage=max(h),design_kappa=kappa(X))
   for(f in c(0,.1,.25,.5,.75,.9,.95)){
    L <- setNames(rep(0,length(coef(fit))),names(coef(fit)));L[pr] <- log2(1-f)
    ct <- if(f==0)data.table(estimate=0,se=0,t=NA_real_,df=df,p=NA_real_,low=0,high=0) else contrast(fit,L)
    nd <- copy(d);nd[[pr]] <- nd[[pr]]+log2(1-f)
    stopifnot(max(abs(predict(fit,nd)-predict(fit,d)-ct$estimate))<1e-8)
    scenarios[[paste(key,pr,f)]] <- cbind(data.table(grouping=g,phase=ph,specification=spec,group=pr,reduction=f,
      treated_below_range=sum(d[condition=="treatment"][[pr]]+log2(1-f)<min(d[[pr]]))),ct)
   }
  }
  for(pair in comparisons){
   # At 50% reduction, improvement A minus improvement B = beta_B - beta_A.
   L <- setNames(rep(0,length(coef(fit))),names(coef(fit)));L[pair[1]] <- -1;L[pair[2]] <- 1
   ct <- contrast(fit,L);ct3 <- contrast(fit,L,V3)
   ct[,`:=`(grouping=g,phase=ph,specification=spec,comparison=paste(pair,collapse="_minus_"),
     HC3_se=ct3$se,HC3_p=ct3$p,HC3_low=ct3$low,HC3_high=ct3$high)]
   contrasts[[paste(key,paste(pair,collapse="/"))]] <- ct
   # Independent prediction-based contrast.
   da <- copy(d);db <- copy(d);da[[pair[1]]] <- da[[pair[1]]]-1;db[[pair[2]]] <- db[[pair[2]]]-1
   stopifnot(max(abs(predict(fit,da)-predict(fit,db)-ct$estimate))<1e-8)
   for(cy in 1:3){
    ff <- lm(form,d[cycle!=cy]);xx <- model.matrix(ff);ok <- ff$rank==ncol(xx)
    deletions[[paste(key,paste(pair,collapse="/"),cy)]] <- data.table(grouping=g,phase=ph,specification=spec,
      comparison=paste(pair,collapse="_minus_"),dropped_cycle=cy,full_rank=ok,residual_df=df.residual(ff),
      estimate=if(ok)sum(L*coef(ff)) else NA_real_)
   }
  }
 }
}
c <- rbindlist(contrasts);stopifnot(nrow(c)==10)
c[,`:=`(BH10=p.adjust(p,"BH",n=10),HC3_BH10=p.adjust(HC3_p,"BH",n=10))]
fwrite(c,file.path(out,"tables/joint_group_50_percent_contrasts.tsv"),sep="\t")
fwrite(rbindlist(audit),file.path(out,"tables/joint_design_audit.tsv"),sep="\t")
fwrite(rbindlist(scenarios),file.path(out,"tables/coordinated_group_reduction_scenarios.tsv"),sep="\t")
fwrite(rbindlist(deletions),file.path(out,"tables/leave_cycle_out_group_contrasts.tsv"),sep="\t")
fwrite(rbindlist(diagnostics),file.path(out,"tables/joint_diagnostics.tsv"),sep="\t")
fwrite(rbindlist(observations,fill=TRUE),file.path(out,"tables/joint_model_observations.tsv"),sep="\t")
saveRDS(fits,file.path(out,"joint_fits.rds"))
print(c[,.(grouping,phase,specification,comparison,estimate,low,high,p,BH10)])
```

```{r}
# All observations retained; asinh scale shows unstable tails without truncating them.
s[,specification:=factor(specification,levels=c("categorical_cycle","linear_cycle","condition_only"),
 labels=c("Categorical cycle + condition","Linear cycle + condition","Condition only"))]
labels <- c(up14="Original 14",rest334="Remaining 334",up="Up (14)",down="Down only (22)",ns="No main-effect hit (248)",not_tested="No tested genes (64)")
colours <- c(up14="#C55A36",rest334="#477CA8",up="#C55A36",down="#477CA8",ns="#8573A8",not_tested="#8A8A8A")
for(g in c("binary","direction")){
 dt <- copy(s);dt[,group:=get(g)]
 lev <- if(g=="binary")c("up14","rest334") else c("up","down","ns","not_tested")
 dt[,group:=factor(group,levels=lev)]
 p <- ggplot(dt,aes(delta_score,colour=group))+stat_ecdf(geom="step",linewidth=.65)+
  geom_vline(xintercept=0,colour="grey60",linetype=2)+
  facet_grid(phase~specification)+scale_colour_manual(values=colours,labels=labels,drop=FALSE)+
  scale_x_continuous(transform="asinh",breaks=c(-20,-5,-1,0,1,5,20))+
  theme_bw(base_size=9)+theme(legend.position="bottom",legend.title=element_blank())+
  labs(x="Predicted score change at 50% lower abundance of one MAG (asinh axis)",y="Cumulative fraction of MAGs")
 for(ext in c("pdf","png"))ggsave(file.path(out,"figures",paste0("individual_MAG_",g,".",ext)),p,width=10,height=5,
    device=if(ext=="pdf")grDevices::pdf else ragg::agg_png,dpi=180)
}
cc <- copy(c);cc[,label:=paste(phase,specification,comparison,sep=" / ")]
p <- ggplot(cc,aes(estimate,reorder(label,estimate)))+geom_vline(xintercept=0,colour="grey60",linetype=2)+
 geom_segment(aes(x=low,xend=high,yend=label),linewidth=.5)+geom_point(size=2)+
 scale_x_continuous(transform="asinh",breaks=c(-100,-20,-5,-1,0,1,5,20,100))+
 theme_bw(base_size=9)+labs(x="Difference in predicted gain at 50% coordinated group reduction (asinh axis)",y=NULL)
for(ext in c("pdf","png"))ggsave(file.path(out,"figures",paste0("joint_group_contrasts.",ext)),p,width=11,height=5,
 device=if(ext=="pdf")grDevices::pdf else ragg::agg_png,dpi=180)
stopifnot(identical(before,vapply(paths,sha,character(1))))
nb <- Sys.getenv("PHAGE_UV_NOTEBOOK_PATH");file.copy(nb,file.path(out,"source",basename(nb)))
fwrite(data.table(path=c(paths,nb),sha256=vapply(c(paths,nb),sha,character(1))),file.path(out,"input_sha256.tsv"),sep="\t")
capture.output(sessionInfo(),file=file.path(out,"sessionInfo.txt"))
writeLines(c("PASS: complete 348-MAG partition, 14/334 binary and 14/22/248/64 directional evidence.",
 "PASS: six unique outcomes per joint fit; saturated designs excluded from inference.",
 "PASS: ten planned joint contrasts; BH10 and HC3 BH10; independent prediction contrasts.",
 "PASS: zero-reduction and graded-scenario prediction contrasts; input hashes unchanged.",
 "Individual-MAG distribution summaries are descriptive, not independent-replicate group tests."),file.path(out,"CHECKS.txt"))
```
