---
title: "All-MAG abundance associations with cleaning performance"
format: html
execute:
  enabled: false
---

## Fixed analysis specification

Extend the existing DNA models to all 348 MAGs without screening on expression
or cleaning-score results. Outcome and normalization remain frozen. Fit each
phase separately to the same six unique membrane-cycle outcomes.
Primary adjustment: categorical cycle and condition. Sensitivities: linear
cycle and condition, and condition only. Residual degrees of freedom are 1, 2
and 3 respectively. Report all models, two-sided OLS and HC3 t approximations,
BH across 696 MAG-phase tests per specification and across all 2088 tests.

Suppression uses log2(1-f) at frozen normalization for f=0,10,25,50,75,90,95%.
Compare score changes at the same fraction. Each single-MAG scenario holds
other modeled covariates fixed; effects from separate models cannot be added.
The slope p-value is the same hypothesis as a nonzero scenario effect, so
suppression fractions do not create independent tests. Pointwise intervals,
extrapolation flags and leave-cycle-out slopes accompany every model.

Expression labels are metadata and do not affect fitting.
Taxonomy is metadata only. Existing 14-candidate DNA results must reproduce.

## Limitations

Condition is confounded with membrane. Model tests assume Gaussian independent
errors conditional on fixed effects, despite repeated observations. These are
exploratory conditional associations. Scenarios are changes in the constructed
index, not demonstrated reductions in attached biofilm or phage efficacy.
Source fraction and omics-cycle alignment assumptions remain as documented in
the original cleaning-score audit. Score-construction uncertainty is unmodeled.

```{r}
suppressPackageStartupMessages({library(data.table);library(ggplot2)})
out <- commandArgs(TRUE)[1];stopifnot(length(out)==1,!is.na(out),!dir.exists(out))
root <- "/Users/shaman.narayanasamy/Work/data/phage_uv_treatment/PRJEB79569/derived"
paths <- c(score=file.path(root,"cleaning_performance_score_2026-09-14-v1/cleaning_score.rds"),
 metadata=file.path(root,"community_cycle_models_2026-09-13-v1/tables/sample_metadata.tsv"),
 dna=file.path(root,"community_cycle_models_2026-09-13-v1/tables/mag_normalized_counts.tsv"),
 raw=file.path(root,"community_cycle_models_2026-09-13-v1/tables/mag_counts.tsv"),
 sf=file.path(root,"community_cycle_models_2026-09-13-v1/tables/mag_size_factors.tsv"),
 membership=file.path(root,"mag_functional_concentration_combined_2026-09-13-v1/tables/membership_and_original_statistics.tsv.gz"))
sha <- function(p)digest::digest(file=p,algo="sha256")
before <- vapply(paths,sha,character(1))
score <- readRDS(paths['score'])$scores
stopifnot(nrow(score)==6,!anyDuplicated(score[,.(condition,cycle)]),
          all(score$published_round==score$cycle+1L))
members <- fread(paths['membership'])[set_id=="biofilm_combined"]
dna <- fread(paths['dna']);raw <- fread(paths['raw']);sf <- fread(paths['sf'])
stopifnot(identical(dna$feature_id,raw$feature_id))
for(sn in sf$sample_title)stopifnot(max(abs(dna[[sn]]-raw[[sn]]/sf[sample_title==sn,size_factor]))<1e-6)
candidates <- sort(dna$feature_id)
stopifnot(length(candidates)==348,!anyDuplicated(candidates))
long <- melt(dna,id.vars="feature_id",variable.name="sample_title",value.name="DNA_count")
setnames(long,"feature_id","MAG_ID")
metadata <- fread(paths['metadata'])
z <- merge(long,metadata[,.(sample_title,condition,phase,cycle)],by="sample_title")
tax <- unique(members[,.(MAG_ID,genus)])
stopifnot(!anyDuplicated(tax$MAG_ID))
z <- merge(z,tax,by="MAG_ID",all.x=TRUE)
z[is.na(genus)|genus=="",genus:="Unassigned"]
stopifnot(nrow(z)==348*12,!anyNA(z$DNA_count),all(z$DNA_count>0))
z[,`:=`(x=log2(DNA_count),predictor="DNA")]
z <- merge(z,score[,.(condition,cycle,published_round,score)],by=c("condition","cycle"),allow.cartesian=TRUE)
stopifnot(nrow(z)==4176,!anyDuplicated(z[,.(MAG_ID,predictor,phase,condition,cycle)]))

```

```{r}
#| label: fits-and-independent-checks
results <- list();obs <- list();deletions <- list();fits <- list();scenarios <- list()
for(spec in c("categorical_cycle","linear_cycle","condition_only"))for(mag in candidates)
 for(pr in "DNA")for(ph in c("initial","backflush")){
 d <- copy(z[MAG_ID==mag & predictor==pr & phase==ph]);setorder(d,condition,cycle)
 d[,condition:=factor(condition,levels=c("control","treatment"))]
 form <- if(spec=="categorical_cycle")score~factor(cycle)+condition+x else if(spec=="linear_cycle")score~cycle+condition+x else score~condition+x
 nf <- if(spec=="categorical_cycle")x~factor(cycle)+condition else if(spec=="linear_cycle")x~cycle+condition else x~condition
 yf <- if(spec=="categorical_cycle")score~factor(cycle)+condition else if(spec=="linear_cycle")score~cycle+condition else score~condition
 fit <- lm(form,d);X <- model.matrix(fit);df <- df.residual(fit)
 stopifnot(nrow(d)==6,fit$rank==ncol(X),df==switch(spec,categorical_cycle=1,linear_cycle=2,condition_only=3))
 b <- unname(coef(fit)['x']);se <- sqrt(vcov(fit)['x','x']);tt <- b/se
 h <- hatvalues(fit);V <- solve(crossprod(X))
 stopifnot(all(h<1-1e-10))
 hc3 <- V%*%crossprod(X,X*as.numeric((resid(fit)/(1-h))^2))%*%V
 se3 <- sqrt(hc3['x','x'])
 rx <- resid(lm(nf,d));ry <- resid(lm(yf,d))
 fwl <- sum(rx*ry)/sum(rx^2)
 stopifnot(abs(fwl-b)<1e-8*max(1,abs(b)))
 key <- paste(spec,mag,pr,ph,sep="/");fits[[key]] <- fit
 dl <- rbindlist(lapply(1:3,function(cy){
  dd<-d[cycle!=cy];ff<-lm(form,dd);xx<-model.matrix(ff)
  data.table(dropped_cycle=cy,beta=if(ff$rank==ncol(xx))unname(coef(ff)['x']) else NA_real_,
             residual_df=df.residual(ff),full_rank=ff$rank==ncol(xx))
 }))
 dl[,`:=`(specification=spec,MAG_ID=mag,predictor=pr,phase=ph,same_sign=sign(beta)==sign(b))]
 deletions[[key]] <- dl
 results[[key]] <- data.table(specification=spec,MAG_ID=mag,genus=d$genus[1],predictor=pr,phase=ph,
 n=6L,residual_df=df,beta=b,se=se,t=tt,p=2*pt(-abs(tt),df),
 CI_low=b-qt(.975,df)*se,CI_high=b+qt(.975,df)*se,
 HC3_se=se3,HC3_p=2*pt(-abs(b/se3),df),
 HC3_low=b-qt(.975,df)*se3,HC3_high=b+qt(.975,df)*se3,
 beta_per_predictor_SD=b*sd(d$x),partial_R2=tt^2/(tt^2+df),
 predictor_VIF=1/(1-summary(lm(nf,d))$r.squared),max_leverage=max(h),
 max_Cook=max(cooks.distance(fit)),design_kappa=kappa(X),
 leave_cycle_same_sign=sum(dl$same_sign,na.rm=TRUE),leave_cycle_estimable=sum(dl$full_rank))
 d[,`:=`(specification=spec,fitted=fitted(fit),residual=resid(fit),partial_x=rx,partial_score=ry)]
 obs[[key]] <- d
 if(pr=="DNA"){
  sc <- data.table(suppression=c(0,.1,.25,.5,.75,.9,.95))
  sc[,delta_x:=log2(1-suppression)]
  sc[,`:=`(specification=spec,MAG_ID=mag,genus=d$genus[1],phase=ph,
    delta_score=b*delta_x,low=(b+qt(.975,df)*se)*delta_x,high=(b-qt(.975,df)*se)*delta_x)]
  sc[,treated_below_observed_range:=vapply(delta_x,function(dx)sum(d[condition=="treatment",x]+dx<min(d$x)),integer(1))]
  scenarios[[key]] <- sc
  changed <- copy(d);changed[,x:=x-1]
  stopifnot(max(abs(predict(fit,changed)-predict(fit,d)+b))<1e-8*max(1,abs(b)))
 }
}
a <- rbindlist(results);stopifnot(nrow(a)==2088)
a[,`:=`(FDR_696=p.adjust(p,"BH",n=696),HC3_FDR_696=p.adjust(HC3_p,"BH",n=696)),by=specification]
a[,`:=`(FDR_2088=p.adjust(p,"BH",n=2088),HC3_FDR_2088=p.adjust(HC3_p,"BH",n=2088))]
sc <- merge(rbindlist(scenarios),a[predictor=="DNA",.(specification,MAG_ID,phase,p,FDR_696,FDR_2088)],
            by=c("specification","MAG_ID","phase"))
stopifnot(nrow(sc)==14616,all(sc[suppression==0,delta_score]==0),all(sc$low<=sc$high))
stopifnot(identical(before,vapply(paths,sha,character(1))))
for(dir in c("tables","figures","source"))dir.create(file.path(out,dir),recursive=TRUE)
fwrite(a,file.path(out,"tables/models.tsv"),sep="\t")
fwrite(rbindlist(obs),file.path(out,"tables/model_observations.tsv"),sep="\t")
fwrite(rbindlist(deletions),file.path(out,"tables/leave_cycle_out.tsv"),sep="\t")
fwrite(sc,file.path(out,"tables/DNA_suppression_scenarios.tsv"),sep="\t")
saveRDS(fits,file.path(out,"fitted_models.rds"))
paths <- c(paths,notebook=Sys.getenv("PHAGE_UV_NOTEBOOK_PATH"))
fwrite(data.table(path=paths,sha256=vapply(paths,sha,character(1))),file.path(out,"input_sha256.tsv"),sep="\t")
file.copy(paths['notebook'],file.path(out,"source",basename(paths['notebook'])))
capture.output(sessionInfo(),file=file.path(out,"sessionInfo.txt"))
writeLines(c("PASS:2088 full-rank six-observation models; residual df1/2/3.",
 "PASS:independent FWL slopes, raw/size-factor reconciliation, fixed candidate membership.",
 "PASS:14616 DNA scenarios, zero-suppression and independent predict contrasts, unchanged source hashes.",
 "Inference conditional on Gaussian independent errors, selected candidates and fixed constructed score.",
 "No calibrated repeated-membrane causal inference. No RNA percentage suppression."),file.path(out,"CHECKS.txt"))
print(a[,.(tests=.N,nominal=sum(p<.05),BH696=sum(FDR_696<.05),BH2088=sum(FDR_2088<.05),
 HC3_BH696=sum(HC3_FDR_696<.05),min_p=min(p),min_FDR=min(FDR_696)),by=specification])
print(a[order(p),.(specification,MAG_ID,genus,predictor,phase,beta,p,FDR_696)][1:12])
```
