---
title: "MAG abundance and expression associations with cleaning performance"
format: html
editor: source
---

## Specification fixed before execution

Outcome: frozen six-row standardized composite from `cleaning_performance_score.qmd`.
The 14 MAGs carrying at least one upregulated combined3905 gene are the fixed
candidate set; candidate selection is post hoc, not independent validation.
DNA predictor: log2 whole-MAG DESeq2-size-factor-normalized count. RNA predictor:
mean gene-centred log2CPM of ALL combined3905 tested genes in each MAG.
Initial and Backflush are separate predictor views, never pooled or used to
duplicate outcomes within one fit. Each model has six observations.

Primary: `lm(score ~ factor(cycle) + condition + x)` (5 parameters, 1 residual df).
Sensitivity: `lm(score ~ cycle + condition + x)` (4 parameters, 2 residual df).
The latter assumes a linear cycle trend. Neither is selected using its p-value.
No condition interaction: categorical-cycle interaction model is saturated.
No random membrane effect: only two membranes, each confounded with condition.

Primary family: 14 MAGs x 2 predictor types x 2 phases = 56 slope tests.
Sensitivity family: another 56. Report two-sided OLS t tests and BH over 56
within specification, plus BH over all 112. HC3 t approximations corrected
separately with the same family sizes. They are conditional model-based
summaries, not calibrated repeated-membrane or post-selection inference.
Gaussian independent errors and fixed predictors/score construction are working
assumptions; digitisation and estimated score-scaling uncertainty are not in CIs.
Deletion of each cycle is an influence check, not predictive validation; saturated
deletion fits have no p-values. No stochastic bootstrap is performed here.

DNA suppression scenarios use delta_x=log2(1-f) with frozen normalization;
score change=beta*delta_x, in index units, NOT percent biofilm reduction.
Do not sum separate MAG effects. RNA slopes are per score unit, not percent
transcript suppression. A negative slope predicts improvement when x is reduced.
The sequenced detached population is not a measured attached-cell population.
Published rounds 2–4 to omics cycles 1–3 remain an explicit order-based assumption;
pooled-phenotype versus third-backflush sequencing ambiguity remains unresolved.

```{r}
suppressPackageStartupMessages({library(data.table);library(ggplot2)})
out <- commandArgs(TRUE)[1];stopifnot(length(out)==1,!is.na(out),!dir.exists(out))
root <- "/Users/shaman.narayanasamy/Work/data/phage_uv_treatment/PRJEB79569/derived"
paths <- c(score=file.path(root,"cleaning_performance_score_2026-09-14-v1/cleaning_score.rds"),
 overlay=file.path(root,"figure_derived_biofilm_phenotypes_2026-09-13-v1/tables/expression_phenotype_overlay.tsv"),
 dna=file.path(root,"community_cycle_models_2026-09-13-v1/tables/mag_normalized_counts.tsv"),
 raw=file.path(root,"community_cycle_models_2026-09-13-v1/tables/mag_counts.tsv"),
 sf=file.path(root,"community_cycle_models_2026-09-13-v1/tables/mag_size_factors.tsv"),
 membership=file.path(root,"mag_functional_concentration_combined_2026-09-13-v1/tables/membership_and_original_statistics.tsv.gz"))
sha <- function(p)digest::digest(file=p,algo="sha256")
before <- vapply(paths,sha,character(1))
score <- readRDS(paths['score'])$scores
stopifnot(nrow(score)==6,!anyDuplicated(score[,.(condition,cycle)]),
          all(score$published_round==score$cycle+1L))
z <- fread(paths['overlay'])
z <- unique(z[,.(MAG_ID,genus,sample_title,condition,phase,cycle,expression_score)])
stopifnot(nrow(z)==168,!anyDuplicated(z[,.(MAG_ID,sample_title)]))
members <- fread(paths['membership'])[set_id=="biofilm_combined"]
candidates <- sort(unique(members[FDR<.05 & logFC>0,MAG_ID]))
stopifnot(length(candidates)==14,setequal(candidates,z$MAG_ID))
dna <- fread(paths['dna']);raw <- fread(paths['raw']);sf <- fread(paths['sf'])
stopifnot(identical(dna$feature_id,raw$feature_id))
for(sn in sf$sample_title)stopifnot(max(abs(dna[[sn]]-raw[[sn]]/sf[sample_title==sn,size_factor]))<1e-6)
long <- melt(dna[feature_id %in% candidates],id.vars="feature_id",variable.name="sample_title",value.name="DNA_count")
setnames(long,"feature_id","MAG_ID")
z <- merge(z,long,by=c("MAG_ID","sample_title"),all.x=TRUE)
stopifnot(!anyNA(z$DNA_count),all(z$DNA_count>0))
z[,DNA:=log2(DNA_count)];z[,RNA:=expression_score]
z <- melt(z,id.vars=c("MAG_ID","genus","sample_title","condition","phase","cycle"),
 measure.vars=c("DNA","RNA"),variable.name="predictor",value.name="x")
z <- merge(z,score[,.(condition,cycle,published_round,score)],by=c("condition","cycle"),allow.cartesian=TRUE)
stopifnot(nrow(z)==336,!anyDuplicated(z[,.(MAG_ID,predictor,phase,condition,cycle)]))
```

```{r}
#| label: fits-and-independent-checks
results <- list();obs <- list();deletions <- list();fits <- list();scenarios <- list()
for(spec in c("categorical_cycle","linear_cycle"))for(mag in candidates)
 for(pr in c("DNA","RNA"))for(ph in c("initial","backflush")){
 d <- copy(z[MAG_ID==mag & predictor==pr & phase==ph]);setorder(d,condition,cycle)
 d[,condition:=factor(condition,levels=c("control","treatment"))]
 form <- if(spec=="categorical_cycle")score~factor(cycle)+condition+x else score~cycle+condition+x
 nf <- if(spec=="categorical_cycle")x~factor(cycle)+condition else x~cycle+condition
 yf <- if(spec=="categorical_cycle")score~factor(cycle)+condition else score~cycle+condition
 fit <- lm(form,d);X <- model.matrix(fit);df <- df.residual(fit)
 stopifnot(nrow(d)==6,fit$rank==ncol(X),df==if(spec=="categorical_cycle")1 else 2)
 b <- unname(coef(fit)['x']);se <- sqrt(vcov(fit)['x','x']);tt <- b/se
 h <- hatvalues(fit);V <- solve(crossprod(X))
 stopifnot(all(h<1-1e-10))
 hc3 <- V%*%crossprod(X,X*as.numeric((resid(fit)/(1-h))^2))%*%V
 se3 <- sqrt(hc3['x','x'])
 rx <- resid(lm(nf,d));ry <- resid(lm(yf,d))
 fwl <- sum(rx*ry)/sum(rx^2)
 stopifnot(abs(fwl-b)<1e-8*max(1,abs(b)))
 key <- paste(spec,mag,pr,ph,sep="/");fits[[key]] <- fit
 dl <- rbindlist(lapply(1:3,function(cy){
  dd<-d[cycle!=cy];ff<-lm(form,dd);xx<-model.matrix(ff)
  data.table(dropped_cycle=cy,beta=if(ff$rank==ncol(xx))unname(coef(ff)['x']) else NA_real_,
             residual_df=df.residual(ff),full_rank=ff$rank==ncol(xx))
 }))
 dl[,`:=`(specification=spec,MAG_ID=mag,predictor=pr,phase=ph,same_sign=sign(beta)==sign(b))]
 deletions[[key]] <- dl
 results[[key]] <- data.table(specification=spec,MAG_ID=mag,genus=d$genus[1],predictor=pr,phase=ph,
 n=6L,residual_df=df,beta=b,se=se,t=tt,p=2*pt(-abs(tt),df),
 CI_low=b-qt(.975,df)*se,CI_high=b+qt(.975,df)*se,
 HC3_se=se3,HC3_p=2*pt(-abs(b/se3),df),
 HC3_low=b-qt(.975,df)*se3,HC3_high=b+qt(.975,df)*se3,
 beta_per_predictor_SD=b*sd(d$x),partial_R2=tt^2/(tt^2+df),
 predictor_VIF=1/(1-summary(lm(nf,d))$r.squared),max_leverage=max(h),
 max_Cook=max(cooks.distance(fit)),design_kappa=kappa(X),
 leave_cycle_same_sign=sum(dl$same_sign,na.rm=TRUE),leave_cycle_estimable=sum(dl$full_rank))
 d[,`:=`(specification=spec,fitted=fitted(fit),residual=resid(fit),partial_x=rx,partial_score=ry)]
 obs[[key]] <- d
 if(pr=="DNA"){
  sc <- data.table(suppression=c(0,.1,.25,.5,.75,.9,.95))
  sc[,delta_x:=log2(1-suppression)]
  sc[,`:=`(specification=spec,MAG_ID=mag,genus=d$genus[1],phase=ph,
    delta_score=b*delta_x,low=(b+qt(.975,df)*se)*delta_x,high=(b-qt(.975,df)*se)*delta_x)]
  sc[,treated_below_observed_range:=vapply(delta_x,function(dx)sum(d[condition=="treatment",x]+dx<min(d$x)),integer(1))]
  scenarios[[key]] <- sc
  changed <- copy(d);changed[,x:=x-1]
  stopifnot(max(abs(predict(fit,changed)-predict(fit,d)+b))<1e-8*max(1,abs(b)))
 }
}
a <- rbindlist(results);stopifnot(nrow(a)==112)
a[,`:=`(FDR_56=p.adjust(p,"BH",n=56),HC3_FDR_56=p.adjust(HC3_p,"BH",n=56)),by=specification]
a[,`:=`(FDR_112=p.adjust(p,"BH",n=112),HC3_FDR_112=p.adjust(HC3_p,"BH",n=112))]
sc <- merge(rbindlist(scenarios),a[predictor=="DNA",.(specification,MAG_ID,phase,p,FDR_56,FDR_112)],
            by=c("specification","MAG_ID","phase"))
stopifnot(nrow(sc)==392,all(sc[suppression==0,delta_score]==0),all(sc$low<=sc$high))
stopifnot(identical(before,vapply(paths,sha,character(1))))
for(dir in c("tables","figures","source"))dir.create(file.path(out,dir),recursive=TRUE)
fwrite(a,file.path(out,"tables/models.tsv"),sep="\t")
fwrite(rbindlist(obs),file.path(out,"tables/model_observations.tsv"),sep="\t")
fwrite(rbindlist(deletions),file.path(out,"tables/leave_cycle_out.tsv"),sep="\t")
fwrite(sc,file.path(out,"tables/DNA_suppression_scenarios.tsv"),sep="\t")
saveRDS(fits,file.path(out,"fitted_models.rds"))
paths <- c(paths,notebook=Sys.getenv("PHAGE_UV_NOTEBOOK_PATH"))
fwrite(data.table(path=paths,sha256=vapply(paths,sha,character(1))),file.path(out,"input_sha256.tsv"),sep="\t")
file.copy(paths['notebook'],file.path(out,"source",basename(paths['notebook'])))
capture.output(sessionInfo(),file=file.path(out,"sessionInfo.txt"))
writeLines(c("PASS:112 full-rank six-observation models, primary df1 and sensitivity df2.",
 "PASS:independent FWL slopes, raw/size-factor reconciliation, fixed candidate membership.",
 "PASS:392 DNA scenarios, zero-suppression and independent predict contrasts, unchanged source hashes.",
 "Inference conditional on Gaussian independent errors, selected candidates and fixed constructed score.",
 "No calibrated repeated-membrane causal inference. No RNA percentage suppression."),file.path(out,"CHECKS.txt"))
print(a[,.(tests=.N,nominal=sum(p<.05),BH56=sum(FDR_56<.05),BH112=sum(FDR_112<.05),
 HC3_BH56=sum(HC3_FDR_56<.05),min_p=min(p),min_FDR=min(FDR_56)),by=specification])
print(a[order(p),.(specification,MAG_ID,genus,predictor,phase,beta,p,FDR_56)][1:12])
```

```{r}
#| label: diagnostic-preview
aa <- copy(a[specification=="categorical_cycle"])
aa[,label:=paste(genus,sub("_MAGScoT_cleanbin_","/",MAG_ID))]
g <- ggplot(aa,aes(beta,label))+geom_vline(xintercept=0,colour="grey70")+
 geom_segment(aes(x=CI_low,xend=CI_high,yend=label),linewidth=.4)+geom_point(size=1.8)+
 facet_wrap(~predictor+phase,scales="free_x",ncol=2)+theme_bw(base_size=9)+
 labs(x="Composite-score slope (DNA: per doubling; RNA: per expression-score unit)",y=NULL,
 subtitle="Cycle-factor + condition adjusted; nominal OLS 95% intervals, 1 residual df",
 caption="Exploratory conditional associations. Each panel uses six outcomes; phases are separate fits.")
ggsave(file.path(out,"figures/primary_associations.png"),g,width=13,height=10,dpi=150,device=ragg::agg_png)
```
