---
title: "MAG and vOTU abundance: condition and cycle interaction"
format: html
editor: source
execute:
  enabled: false
---

Run `bash scripts/run_qmd.sh analysis/community_cycle_models.qmd NEW_OUTPUT`.
Append `--catalogue-aligned` to include all 634 high-quality, viral-gene-positive
CheckV representatives, irrespective of Cenote annotation availability. The
default preserves the archived 616-vOTU eligibility for reproducibility.
Uses the same CoverM counts, 348-MAG/616-vOTU catalogue rules and count filter
as `scripts/run_community_differential_abundance.qmd`. Does not overwrite it.
DESeq2 additive Wald test: `~phase + cycle + condition`. Joint interaction LRT:
full `~phase + cycle * condition`, reduced `~phase + cycle + condition` (2 df).
No random effects or autocorrelation model. System-specific exploratory evidence,
not replicated treatment inference. Counts are relative sequencing abundance,
not absolute microbial load. Each assay has a separate normalisation and BH family.

```{r}
#| label: input-contract
suppressPackageStartupMessages({library(data.table);library(DESeq2);library(digest)})
args <- commandArgs(trailingOnly=TRUE)
catalogue_aligned <- "--catalogue-aligned" %in% args
args <- args[args != "--catalogue-aligned"]
stopifnot(length(args)==1L)
out <- args[[1]]
stopifnot(!dir.exists(out))
repo <- normalizePath(getwd())
root <- "/Users/shaman.narayanasamy/Work/data/phage_uv_treatment/PRJEB79569"
vp <- file.path(root,"viromics/annotation/PRJEB79569_vOTUs/vclust_catalogue")
paths <- c(counts=file.path(root,"quantification/mags_votu/coverage/metagenomics/coverm/output-Read_Count.tsv"),
 metadata=file.path(repo,"metadata/sample_metadata.tsv"),
 checkv=file.path(vp,"checkv/quality_summary.tsv"),
 cenote=file.path(vp,"cenotetaker3/output/output_virus_summary.tsv"),
 notebook=Sys.getenv("PHAGE_UV_NOTEBOOK_PATH"))
if(catalogue_aligned) paths <- c(paths, composition_catalogue=file.path(root,
 "derived/votu_taxonomic_resolution_review_2026-09-13-v4/tables/catalogue_rank_assignments.tsv"))
stopifnot(all(file.exists(paths)))
sha <- function(p)digest(file=p,algo="sha256")
hashes<-vapply(paths,sha,character(1))
for(s in c("tables","models","source"))dir.create(file.path(out,s),recursive=TRUE)
fwrite(data.table(name=names(paths),path=paths,sha256=hashes),file.path(out,"input_sha256.tsv"),sep="\t")
m<-fread(paths[["metadata"]])[,.(sample_title,condition,phase,cycle,metagenome_run_accession)]
m[,`:=`(condition=factor(condition,c("control","treatment")),phase=factor(phase,c("initial","backflush")),cycle=factor(cycle,c("1","2","3")))]
setorder(m,cycle,phase,condition)
stopifnot(nrow(m)==12,!anyDuplicated(m$sample_title),all(table(m$condition,m$phase,m$cycle)==1))
md<-as.data.frame(m);rownames(md)<-md$sample_title
dmain<-model.matrix(~phase+cycle+condition,md)
dfull<-model.matrix(~phase+cycle*condition,md)
stopifnot(qr(dmain)$rank==5,qr(dfull)$rank==7)
fwrite(m,file.path(out,"tables/sample_metadata.tsv"),sep="\t")
for(n in c("dmain","dfull"))fwrite(as.data.table(get(n),keep.rownames="sample_title"),file.path(out,"tables",paste0(n,".tsv")),sep="\t")
cv<-fread(paths[["checkv"]],select=c("contig_id","viral_genes","miuvig_quality"))
ct<-fread(paths[["cenote"]],select=c("input_name","virus_seq_length","taxonomy_hierarchy"));setnames(ct,"input_name","contig_id")
v<-merge(ct,cv,by="contig_id",all.x=TRUE);setorder(v,contig_id,-virus_seq_length,taxonomy_hierarchy,na.last=TRUE)
v<-v[,.SD[1L],by=contig_id][miuvig_quality=="High-quality" & viral_genes>0]
stopifnot(nrow(v)==616)
archived_ids <- v$contig_id
if(catalogue_aligned){
 stopifnot(!anyDuplicated(cv$contig_id))
 v <- cv[miuvig_quality=="High-quality" & viral_genes>0]
 catalogue <- fread(paths[["composition_catalogue"]])
 stopifnot(nrow(v)==634L,!anyDuplicated(catalogue$contig_id),
           setequal(v$contig_id,catalogue$contig_id),all(archived_ids %chin% v$contig_id))
}
fwrite(data.table(feature_id=v$contig_id,archived_eligible=v$contig_id %chin% archived_ids,
 catalogue_mode=if(catalogue_aligned)"checkv_catalogue_634" else "archived_cenote_616"),
 file.path(out,"tables/votu_eligibility.tsv"),sep="\t")
x<-fread(paths[["counts"]]);cols<-setdiff(names(x),"Contig")
acc<-sub(".*(ERR[0-9]+).*","\\1",cols)
stopifnot(!anyDuplicated(acc),setequal(acc,m$metagenome_run_accession))
setnames(x,cols,m$sample_title[match(acc,m$metagenome_run_accession)])
samples<-m$sample_title
mg<-x[grepl("_MAGScoT_",Contig,fixed=TRUE)]
mg[,feature_id:=vapply(strsplit(Contig,"_",fixed=TRUE),function(z)paste(z[1:4],collapse="_"),character(1))]
mg<-mg[,lapply(.SD,sum,na.rm=TRUE),by=feature_id,.SDcols=samples]
vv<-x[Contig %chin% v$contig_id,c("Contig",samples),with=FALSE];setnames(vv,"Contig","feature_id")
stopifnot(nrow(mg)==348,nrow(vv)==nrow(v),!anyDuplicated(vv$feature_id),
          setequal(vv$feature_id,v$contig_id))
rm(x);gc()
```

```{r}
#| label: fit-and-export
summary_rows<-list()
for(layer in c("mag","votu")){
 message("Fitting ",layer)
 tab<-if(layer=="mag")mg else vv
 mat<-as.matrix(tab[,..samples]);rownames(mat)<-tab$feature_id
 stopifnot(all(is.finite(mat)),all(mat>=0),max(mat)<.Machine$integer.max)
 rounded<-round(mat);storage.mode(rounded)<-"integer"
 keep<-rowSums(rounded>=10)>=3
 fwrite(data.table(feature_id=rownames(mat),retained=keep,max_rounding_delta=apply(abs(mat-rounded),1,max)),file.path(out,"tables",paste0(layer,"_filter.tsv")),sep="\t")
 fwrite(data.table(feature_id=rownames(rounded),rounded),file.path(out,"tables",paste0(layer,"_counts.tsv")),sep="\t")
 base<-DESeqDataSetFromMatrix(rounded[keep,,drop=FALSE],md,~phase+cycle+condition)
 main<-DESeq(base,fitType="local",quiet=TRUE)
 full<-base;design(full)<-~phase+cycle*condition
 full<-DESeq(full,fitType="local",test="LRT",reduced=~phase+cycle+condition,quiet=TRUE)
 stopifnot(isTRUE(all.equal(sizeFactors(main),sizeFactors(full))))
 # Save fitted objects so a later agent can inspect likelihoods and diagnostics.
 saveRDS(main,file.path(out,"models",paste0(layer,"_main.rds")))
 saveRDS(full,file.path(out,"models",paste0(layer,"_interaction_lrt.rds")))
 a<-as.data.table(as.data.frame(results(main,contrast=c("condition","treatment","control"),alpha=.05)),keep.rownames="feature_id")
 b<-as.data.table(as.data.frame(results(full,alpha=.05)),keep.rownames="feature_id")
 # The LRT's displayed coefficient is NOT its joint effect. Omit it from joint table.
 b[,c("log2FoldChange","lfcSE"):=NULL]
 b[,test_df:=2L]
 stopifnot(all(b$stat[is.finite(b$stat)]>=-1e-8))
 for(z in c("a","b")){
  q<-get(z);q[,padj_BH_all_nonmissing:=p.adjust(pvalue,"BH")]
  fwrite(q,file.path(out,"tables",paste0(layer,if(z=="a")"_main_wald.tsv" else "_interaction_lrt.tsv")),sep="\t")
 }
 # Secondary Wald coefficients use the full model's dispersion estimates.
 w<-nbinomWaldTest(full,quiet=TRUE)
 rn<-resultsNames(w);ints<-grep("cycle[23].conditiontreatment",rn,value=TRUE)
 stopifnot(length(ints)==2)
 co<-rbindlist(lapply(ints,function(n){z<-as.data.table(as.data.frame(results(w,name=n,alpha=.05)),keep.rownames="feature_id");z[,coefficient:=n];z}))
 co[,padj_BH_two_coefficients:=p.adjust(pvalue,"BH")]
 fwrite(co,file.path(out,"tables",paste0(layer,"_interaction_coefficients.tsv")),sep="\t")
 fwrite(data.table(sample_title=samples,size_factor=sizeFactors(main)),file.path(out,"tables",paste0(layer,"_size_factors.tsv")),sep="\t")
 normalized<-counts(main,normalized=TRUE)
 fwrite(data.table(feature_id=rownames(normalized),normalized),file.path(out,"tables",paste0(layer,"_normalized_counts.tsv")),sep="\t")
 for(n in c("main","full")){
  fit<-get(n);dg<-as.data.table(as.data.frame(mcols(fit)),keep.rownames="feature_id")
  fwrite(dg,file.path(out,"tables",paste0(layer,"_",n,"_diagnostics.tsv")),sep="\t")
  fwrite(data.table(feature_id=rownames(fit),assays(fit)[["cooks"]]),file.path(out,"tables",paste0(layer,"_",n,"_cooks.tsv")),sep="\t")
 }
 summary_rows[[layer]]<-data.table(layer=layer,input_features=nrow(mat),retained=sum(keep),
  main_fdr05=sum(a$padj<.05,na.rm=TRUE),interaction_fdr05=sum(b$padj<.05,na.rm=TRUE),
  main_unfiltered_BH05=sum(a$padj_BH_all_nonmissing<.05,na.rm=TRUE),interaction_unfiltered_BH05=sum(b$padj_BH_all_nonmissing<.05,na.rm=TRUE),
  main_nonconverged=sum(!mcols(main)$betaConv),full_nonconverged=sum(!mcols(full)$fullBetaConv),
  residual_df_main=7L,residual_df_full=5L)
}
fwrite(rbindlist(summary_rows),file.path(out,"tables/summary.tsv"),sep="\t")
stopifnot(identical(hashes,vapply(paths,sha,character(1))))
file.copy(paths[["notebook"]],file.path(out,"source/community_cycle_models.qmd"))
writeLines(capture.output(sessionInfo()),file.path(out,"sessionInfo.txt"))
payload<-list.files(out,recursive=TRUE,full.names=TRUE)
fwrite(data.table(path=substring(payload,nchar(out)+2),sha256=vapply(payload,sha,character(1))),file.path(out,"output_sha256.tsv"),sep="\t")
print(rbindlist(summary_rows))
```

DESeq2 references: https://bioconductor.org/packages/release/bioc/vignettes/DESeq2/inst/doc/DESeq2.html
Default independent filtering and Cook's handling are retained for parity with
the additive predecessor. Additional all-nonmissing BH columns expose the effect
of independent filtering; missing p-values are not silently converted to 1.
No model significance is inferred from plotted normalised counts.
