---
title: "MAG-resolved biofilm and SOS concentration: exploratory allocation tests"
format: html
editor: source
---

Run with `bash scripts/run_qmd.sh analysis/mag_functional_concentration.qmd NEW_OUTPUT`.
Frozen full-transcriptome FDR is retained. These are post hoc gene-allocation
tests, not sample permutations or causal treatment tests. Both historical keyword
variants are reported; expanded biofilm is a sensitivity. No publication sign-off
until upstream negative-F diagnostics and genomic-dependence sensitivity are resolved.

Global primary statistic: Simpson concentration of supported genes across MAGs.
Secondary: maximum MAG share, calibrated across all MAGs. Baseline draws select
K of N eligible set genes uniformly without replacement. Matched draws preserve
K in each condition-blind abundance-tertile x length-tertile x prevalence stratum
(prevalence <=6, 7-11, 12). Quantile ties are never split. Sparse strata are not
merged: fully selected strata are fixed and their contribution is reported.
This conservative matching may leave little randomisable information.
Four endpoints: condition unsigned, up, down, joint interaction (unsigned).
BH spans all sets/endpoints separately for each null and statistic. Per-MAG
baseline concentration tails and within-MAG specificity tails have separate BH
families across all sets/endpoints/MAGs. Dependence is NOT solved by matching.

```{r}
#| label: reproducible-inputs
suppressPackageStartupMessages(library(data.table))
out <- commandArgs(TRUE)[1]
combined_mode <- length(commandArgs(TRUE))>=2 && commandArgs(TRUE)[2]=="combined"
stopifnot(!is.na(out), !dir.exists(out))
root <- "/Users/shaman.narayanasamy/Work/data/phage_uv_treatment/PRJEB79569/derived"
src <- c(condition=file.path(root,"full_transcriptome_de/tables/de_condition_adjusted.tsv"),
 joint=file.path(root,"full_transcriptome_de/tables/de_condition_cycle_interaction_omnibus.tsv"),
 biofilm=file.path(root,"biofilm_full_transcriptome_review_2026-09-13-v4/tables/gene_set_membership.tsv"),
 stress=file.path(root,"full_de_interpretation/tables/tested_feature_category_membership.tsv"),
 taxonomy=file.path(root,"full_de_interpretation/tables/full_de_gene_annotation_taxonomy.tsv.gz"),
 notebook=Sys.getenv("PHAGE_UV_NOTEBOOK_PATH"))
stopifnot(all(file.exists(src)))
if(combined_mode) src['biofilm']<-file.path(root,"biofilm_combined_definition_2026-09-13-v1/combined_gene_membership.tsv")
stopifnot(all(file.exists(src)))
hash <- function(p) digest::digest(file=p,algo="sha256")
before <- vapply(src,hash,character(1))
cnd <- fread(src['condition']); joint <- fread(src['joint'])
stopifnot(!anyDuplicated(cnd$feature_id),!anyDuplicated(joint$feature_id),
 setequal(cnd$feature_id,joint$feature_id))
joint <- joint[match(cnd$feature_id,feature_id)]
cnd[,joint_FDR:=joint$FDR]
stopifnot(all(is.finite(cnd$FDR)),all(is.finite(cnd$joint_FDR)))
bio <- fread(src['biofilm'])[,.(feature_id,set_id)]
stress <- fread(src['stress'])[,.(feature_id,set_id=category)]
if(combined_mode) stress<-stress[set_id=="SOS_response"]
membership <- unique(rbind(bio,stress))
stopifnot(all(membership$feature_id %in% cnd$feature_id))
universe <- cnd[!is.na(MAG_ID) & nzchar(MAG_ID)]
membership <- merge(membership,universe,by="feature_id",allow.cartesian=TRUE)
tax <- unique(fread(src['taxonomy'],select=c("feature_id","phylum","family","genus")))
stopifnot(!anyDuplicated(tax$feature_id))
membership <- merge(membership,tax,by="feature_id",all.x=TRUE)
qbin <- function(x) {
 b <- unique(as.numeric(quantile(x,c(0,1/3,2/3,1),na.rm=TRUE)))
 if(length(b)<2) return(rep(1L,length(x)))
 as.integer(cut(x,b,include.lowest=TRUE))
}
stopifnot(all(is.finite(universe$mean_CPM)),all(universe$gene_length>0),
 all(is.finite(universe$gene_length)))
universe[,stratum:=paste(qbin(mean_CPM),qbin(gene_length),
 fifelse(n_samples_detected<=6,1L,fifelse(n_samples_detected<12,2L,3L)),sep="_")]
membership[,stratum:=universe$stratum[match(feature_id,universe$feature_id)]]
for(s in c("tables","source")) dir.create(file.path(out,s),recursive=TRUE)
fwrite(data.table(input=names(src),path=src,sha256=before),file.path(out,"input_sha256.tsv"),sep="\t")
fwrite(data.table(full_tested=nrow(cnd),MAG_assigned=nrow(universe),
 membership_missing_MAG=sum(!unique(rbind(bio,stress))$feature_id %in% universe$feature_id)),
 file.path(out,"tables/universe_audit.tsv"),sep="\t")
fwrite(membership,file.path(out,"tables/membership_and_original_statistics.tsv.gz"),sep="\t")
```

```{r}
#| label: null-engine-and-tests
# Exact multivariate hypergeometric draws via sequential conditional draws.
draw_alloc <- function(n,k,B) {
 stopifnot(k>=0,k<=sum(n),all(n>=0))
 ans <- matrix(0L,B,length(n)); remaining <- rep(k,B); N <- sum(n)
 for(i in seq_along(n)) {
  if(n[i]>0) {
   ans[,i] <- if(N==n[i]) remaining else rhyper(B,n[i],N-n[i],remaining)
   remaining <- remaining-ans[,i]
  }
  N <- N-n[i]
 }
 stopifnot(all(rowSums(ans)==k)); ans
}
RNGkind("L'Ecuyer-CMRG"); set.seed(20260913)
test <- draw_alloc(c(5,15),4,100000)
stopifnot(abs(mean(test[,1])-1)<.02)
set.seed(11); t1<-draw_alloc(c(2,3,5),3,100)
set.seed(11); stopifnot(identical(t1,draw_alloc(c(2,3,5),3,100)),
 all(draw_alloc(c(0,3),0,10)==0),all(draw_alloc(c(2,3),5,10)[,1]==2))
set.seed(20260913)
B <- 100000L
endpoint <- function(z,e) switch(e,condition=z$FDR<.05,
 up=z$FDR<.05 & z$logFC>0,down=z$FDR<.05 & z$logFC<0,
 joint=z$joint_FDR<.05)
global <- list(); per_mag <- list(); strata <- list(); nulls <- list()
counter <- 0L
for(s in sort(unique(membership$set_id))) for(e in c("condition","up","down","joint")) {
 z<-membership[set_id==s]; hit<-endpoint(z,e); K<-sum(hit); N<-nrow(z)
 mags<-sort(unique(z$MAG_ID)); ni<-tabulate(match(z$MAG_ID,mags),length(mags))
 hi<-tabulate(match(z$MAG_ID[hit],mags),length(mags))
 allhit<-endpoint(universe,e)
 mi<-tabulate(match(universe$MAG_ID,mags),length(mags))
 mh<-tabulate(match(universe$MAG_ID[allhit],mags),length(mags))
 per_mag[[length(per_mag)+1L]]<-data.table(set_id=s,endpoint=e,MAG_ID=mags,
  eligible_set=ni,set_hits=hi,eligible_MAG=mi,MAG_all_hits=mh,
  expected_across=K*ni/N,p_across=phyper(hi-1,ni,N-ni,K,lower.tail=FALSE),
  expected_within=mh*ni/mi,p_within=phyper(hi-1,ni,mi-ni,mh,lower.tail=FALSE))
 for(mode in c("baseline","matched")) {
  counter<-counter+1L
  if(K<2) {
   global[[counter]]<-data.table(set_id=s,endpoint=e,null=mode,N=N,K=K,
    status="insufficient_hits",p_simpson=NA_real_,p_max=NA_real_)
   next
  }
  alloc<-matrix(0L,B,length(mags))
  groups<-if(mode=="baseline")list(seq_len(N)) else split(seq_len(N),z$stratum)
  frozen<-0L
  for(j in seq_along(groups)) {
   ix<-groups[[j]]; kg<-sum(hit[ix]); ng<-length(ix)
   if(kg==ng) frozen<-frozen+kg
   counts<-tabulate(match(z$MAG_ID[ix],mags),length(mags))
   active<-which(counts>0)
   alloc[,active]<-alloc[,active,drop=FALSE]+draw_alloc(counts[active],kg,B)
   strata[[length(strata)+1L]]<-data.table(set_id=s,endpoint=e,null=mode,
    stratum=j,N=ng,K=kg,fixed=kg==0 || kg==ng)
  }
  stopifnot(all(rowSums(alloc)==K))
  sim<-rowSums((alloc/K)^2); mx<-apply(alloc,1,max)/K
  obs<-sum((hi/K)^2); obsmax<-max(hi)/K
  exceed<-sum(sim>=obs-1e-12); exceedmax<-sum(mx>=obsmax-1e-12)
  ci<-binom.test(exceed,B)$conf.int
  global[[counter]]<-data.table(set_id=s,endpoint=e,null=mode,N=N,K=K,
   status="exploratory_gene_independence",fixed_hits=frozen,simpson=obs,max_share=obsmax,
   null_mean_simpson=mean(sim),p_simpson=(exceed+1)/(B+1),
   p_simpson_MC_low=ci[1],p_simpson_MC_high=ci[2],p_max=(exceedmax+1)/(B+1),B=B)
  nulls[[paste(s,e,mode,sep="/")]]<-list(simpson=sim,max_share=mx,expected_MAG=colMeans(alloc),MAG_ID=mags)
 }
 message(s," / ",e,": ",K," supported genes")
}
```

```{r}
#| label: auditable-outputs
g<-rbindlist(global,fill=TRUE)
g[,FDR_simpson:=p.adjust(p_simpson,"BH"),by=null]
g[,FDR_max:=p.adjust(p_max,"BH"),by=null]
p<-rbindlist(per_mag)
p[,FDR_across:=p.adjust(p_across,"BH")]
p[,FDR_within:=p.adjust(p_within,"BH")]
fwrite(g,file.path(out,"tables/global_concentration.tsv"),sep="\t")
fwrite(p,file.path(out,"tables/MAG_specificity_baselines.tsv"),sep="\t")
fwrite(rbindlist(strata),file.path(out,"tables/matching_strata.tsv"),sep="\t")
saveRDS(nulls,file.path(out,"null_distributions.rds"))
capture.output(sessionInfo(),file=file.path(out,"sessionInfo.txt"))
writeLines(c("Seed: 20260913; RNG: L'Ecuyer-CMRG; B: 100000", 
 "Engine tests PASS: K conservation, theoretical mean, reproducibility, edge cases.",
 "Pending: within-MAG matched null, genomic-block sensitivity, synthetic dependence tests, upstream negative-F resolution.",
 "Exploratory only; no manuscript or figure changes."),file.path(out,"CHECKS.txt"))
stopifnot(identical(before,vapply(src,hash,character(1))))
file.copy(src['notebook'],file.path(out,"source",basename(src['notebook'])))
files<-list.files(out,recursive=TRUE,full.names=TRUE)
fwrite(data.table(path=substring(files,nchar(out)+2),sha256=vapply(files,hash,character(1))),
 file.path(out,"output_sha256.tsv"),sep="\t")
```
