---
title: "PRJEB79569 verified manuscript analysis"
format:
  html:
    toc: true
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
---

## Biofilm review addendum

The restored biofilm-associated transcription analysis is maintained in
[biofilm_full_transcriptome_review.qmd](biofilm_full_transcriptome_review.qmd).
Its versioned v4 output is registered as `biofilm_postmodel_review_v4` in the data
manifest. See [the review handoff](../notes/biofilm-review-2026-09-13.md) for results,
figure legends, sensitivity checks and the remaining upstream numerical QC gate.
This is a local author-review candidate, not an adopted manuscript replacement.

```{r}
library(data.table)
library(knitr)

repo_root <- normalizePath("..")
data_manifest <- fread(file.path(repo_root, "manifests", "data_manifest.tsv"))
analysis_registry <- fread(file.path(repo_root, "manuscript", "analysis_registry.tsv"))
claim_registry <- fread(file.path(repo_root, "manuscript", "claim_evidence_registry.tsv"))

manifest_path <- function(id) {
  hit <- data_manifest[data_id == id]
  stopifnot(nrow(hit) == 1L)
  path <- hit$path[[1L]]
  if (!grepl("^/", path)) path <- file.path(repo_root, path)
  normalizePath(path, mustWork = TRUE)
}

full_de_root <- manifest_path("full_transcriptome_de_results")
interpretation_root <- manifest_path("full_de_interpretation")
candidate_root <- manifest_path("mag_genomic_variation_dossiers")
```

## Interpretation boundary

This report reads only verified, manifest-registered outputs. It does not refit a
subset-restricted model. Condition is confounded with membrane identity, so its
coefficient describes the persistent difference between the treated and control
membrane systems. SOS transcription is not direct evidence of DNA damage,
mutagenesis, or adaptation. The 16S stream is separately owned, and host-phage
integration remains deferred.

## Analysis registry

```{r}
kable(analysis_registry)
```

## Complete transcriptome-wide differential expression

```{r}
filtering <- fread(file.path(full_de_root, "tables", "filtering_summary.tsv"))
audit <- fread(file.path(full_de_root, "tables", "deterministic_result_summary.tsv"))
kable(filtering)
kable(audit)
```

```{r}
condition_pdf <- file.path(full_de_root, "figures", "condition-effect-landscape.pdf")
if (file.exists(condition_pdf)) knitr::include_graphics(condition_pdf)
```

## Frozen functional categories

```{r}
category_results <- fread(file.path(
  interpretation_root,
  "tables",
  "functional_category_enrichment.tsv"
))
kable(category_results)
```

```{r}
knitr::include_graphics(file.path(
  interpretation_root,
  "figures",
  "functional-category-enrichment.pdf"
))
```

## Organism-resolved coherence

```{r}
mag_results <- fread(file.path(
  interpretation_root,
  "tables",
  "mag_coherence.tsv"
))
mag_summary <- mag_results[eligible == TRUE, .(
  eligible_MAGs = uniqueN(MAG_ID),
  supported_MAGs = sum(FDR < 0.05),
  higher_ranked = sum(FDR < 0.05 & Direction == "Up"),
  lower_ranked = sum(FDR < 0.05 & Direction == "Down")
), by = coefficient]
kable(mag_summary)
```

```{r}
knitr::include_graphics(file.path(
  interpretation_root,
  "figures",
  "mag-condition-coherence.pdf"
))
```

## Six-cell descriptive recurrence

```{r}
gene_candidates <- fread(file.path(
  interpretation_root,
  "tables",
  "predeclared_gene_candidates.tsv"
))
recurrence_summary <- gene_candidates[, .(features = .N), by = recurrence_class]
kable(recurrence_summary)
```

```{r}
knitr::include_graphics(file.path(
  interpretation_root,
  "figures",
  "recurrent-gene-cell-effects.pdf"
))
```

## Descriptive population-genomic context

```{r}
popgen_qc <- file.path(
  candidate_root,
  "tables",
  "population_genomics_mag_qc.tsv"
)
if (file.exists(popgen_qc)) kable(fread(popgen_qc))
```

## Claim audit

Figure numbering and main-versus-supplement allocation remain deliberately
unassigned.

```{r}
kable(claim_registry)
```
