---
title: "Phage-UV ecology and strain-level adaptation"
author: "Shaman Narayanasamy"
format:
  html:
    toc: true
execute:
  echo: true
  warning: false
  message: false
params:
  sample_metadata: "../metadata/sample_metadata.tsv"
  data_manifest: "../manifests/data_manifest.tsv"
  code_manifest: "../manifests/code_manifest.tsv"
  uv_signatures: "../resources/uv_resistance_signatures.tsv"
---

> **Quarantined legacy notebook.** This scaffold encodes the retired
> SOS-first, DNA-damage, adaptation, and treatment-driven SNV framing. It is
> retained as provenance only. Use `analysis/phage_uv_ecology.qmd` for the
> verified #22/#23 manuscript workflow.

## Purpose

This Quarto document is the downstream manuscript scaffold. It is intentionally manifest-first: heavy upstream outputs are staged on Isilon/HPC and consumed here only after they are listed in `manifests/data_manifest.tsv`.

```{r setup}
library(tidyverse)
```

## Load project metadata

```{r load_metadata}
sample_metadata <- read_tsv(params$sample_metadata, show_col_types = FALSE)
data_manifest <- read_tsv(params$data_manifest, show_col_types = FALSE)
code_manifest <- read_tsv(params$code_manifest, show_col_types = FALSE)
uv_signatures <- read_tsv(params$uv_signatures, show_col_types = FALSE)

sample_metadata %>%
  count(condition, phase, cycle, analysis_group)
```

## Analysis grouping

Downstream manuscript figures use `analysis_group = condition_cycle`. Initial and backflush samples remain traceable through `phase`, but are not treated as separate biological axes in the main figures.

```{r grouping_summary}
sample_metadata %>%
  group_by(analysis_group, condition, cycle) %>%
  summarise(
    phases = paste(sort(unique(phase)), collapse = ";"),
    metagenome_runs = paste(metagenome_run_accession, collapse = ";"),
    n_physical_samples = n(),
    .groups = "drop"
  )
```

## UV signature resource

```{r uv_signature_summary}
uv_signatures %>%
  count(tier, category, specificity)
```

## Staging checklist

```{r staging_checklist}
data_manifest %>%
  count(status, storage_tier) %>%
  arrange(status, storage_tier)
```

## Next executable sections

Add sections here once the staged files exist:

- MAG/rMAG abundance and taxonomy overview.
- UV/DNA-damage signature genome-potential mapping.
- UV/DNA-damage metatranscriptomic activity by `analysis_group`.
- vOTU and CRISPR phage-host link integration.
- inStrain SNV divergence summaries.
- MGnify concordance, only if MGnify analyses are available and mappable.
